Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-12-07 analysis/paper/paper.Rmd
first draft methods complete

Our goal was to estimate the abundance of potentially piscivorous predators in the Deadwater Slough using a mark-recapture approach both during the fall and spring near the peaks of the juvenile Chinook Salmon DSR and NRR emigrations, respectively. The initial intent was to sample during fall 2019 and spring 2020 to estimate predator abundances during the emigration of the 2018 brood year. However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 sampling event was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional sampling event was added for spring 2021. The study design was such that the fall 2019 and fall 2020 sampling efforts included both a "mark week" and "recapture week" intended for mark-recapture models. However, due to budget constraints the spring 2021 effort only included a single "mark week"; instead, we calculated the catch-per-unit-effort (CPUE; fish per hour) for the spring 2021 sampling event for comparison to the fall efforts. In the end, sampling dates were as follows: fall 2019 (mark = November 12-14; recapture = November 19-21), fall 2020 (mark = October 20-23; recapture = October 27-30); spring 2021 (mark = May 18-21). The recapture events were held just one week following the mark events to reduce opportunity for immigration/emigration of predators between events i.e., to avoid violation of a closed population for mark-recapture models.

We attempted a multiple of capture methods during the first year including raft electrofishing, fyke netting, snorkeling, and angling to reduce selectivity, both for species and size classes. Unfortunately, some of the methods proved ineffective. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. And due to debris floating in the water column, there were concerns with fyke nets being pulled from their anchored locations and becoming entangled. Additionally, there were concerns with raft electrofishers and fyke nets incidentally capturing adult steelhead that are present in the mainstem Salmon River during fall and spring. Eventually, angling proved to be the most effective method for capturing piscivorous predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were allowed to fish anywhere within Deadwater Slough; periodically or upon filling a livewell anglers would boat or hike their catch to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, processors would record the date of capture, species, total length (TL; mm), and whether the fish was previously marked. If not previously marked, each fish was provided a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique for each marking day. We also recorded whether marked individuals were released back to be captured again. For release, processors attempted to distribute released individuals throughout Deadwater Slough to facilitate mixing back into the population. We also recorded the crew (person or combination of persons) that captured each fish and the start and end times for each crew were recorded to estimate total effort.

@@

1. The population in closed (no immigration, emigration, births or deaths during the sampling period)
1. Marking a fish does not affect its chances of being caught
1. No loss of marks
1. No marks are missed or mistaken
1. All fish have the same chance of being caught in the second (and subsequent) sample(s)

The estimators fell into two broad categories: single census and multiple census. For the single census estimators, we treated the first week of sampling as the mark event, and the following (second) week as the recapture event, pooling data within each of those two weeks. Alternatively, the multiple census estimators treat each day as a survey, and use information about the total marked fish from all previous surveys to infer the total abundance.

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 2021 and released fish were released unmarked. Instead, to estimate the abundance of Northern Pikeminnow in the spring, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the abundance of Northern Pikeminnow in the fall. This assumes the same capture probability in the fall as in the spring. If that assumption is not true and capture probability did differ, we believe that the capture probability to be lower in the spring due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we consider this a conservative estimate of the number of Northern Pikeminnow present in the Deadwater Slough during the spring emigration of NRR Chinook Salmon.

@@

Lastly, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon, other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner, etc.), and non-targets (e.g., macroinvertebrates, organic matter, etc.). Stomach contents were preserved with 99% isopropyl alcohol in whirl-paks and analyzed in a laboratory. For each sample, wet weight (grams) was recorded for the combined stomach content and for the fish content. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknown. A subset of Northern Pikeminnow captures were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the consumption of juvenile Chinook Salmon outmigrants by Northern Pikeminnow in the Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RcitationFromNick]. The daily rate of consumption in grams for an individual Northern Pikeminnow was based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight, converted from the average length of Northern Pikeminnow captured during our study within the FSA package [@Ogle2021]. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight of Northern Pikeminnow.

We ran three bioenergetic models for this study. For the first model, we chose a 78-day period from September 1 through November 17 where DSR emigrants are known to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin their downstream migration. During this time, water temperatures exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon. This model assumed no growth for Northern Pikeminnow. The second model was run for one full year and assumed no growth for Northern Pikeminnow. The assumption of no growth for pikeminnow was included to show what may be occurring if the population of Northern Pikeminnow is stable. If growth of the population is occurring, the alternative model shows how an increase of 10% body weight changes the estimates of consumption over a full year period. To estimate the number of possible Chinook Salmon that could be consumed, we assumed a weight of 12 g per individual Chinook Salmon. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

>>>>

The first objective was to estimate the population size of piscine predators in the Deadwater Slough to begin to their potential impacts on juvenile Chinook Salmon. Our initial intent was to estimate predator abundances using mark-recapture methods during fall 2019 and spring 2020 during the emigration of the 2018 brood year. We attempted to sample near the peaks of the juvenile DSR (fall) and NRR (spring) emigrations. However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 sampling event was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional sampling event was added for spring 2021. The study design was such that the fall 2019 and fall 2020 sampling efforts included both a "mark week" and "recapture week" intended for mark-recapture models. However, due to budget constraints the spring 2021 effort only included a single "mark week"; instead, we calculated the catch-per-unit-effort (CPUE; fish per hour) for the spring 2021 sampling event for comparison to the fall efforts. In the end, sampling dates were as follows: fall 2019 (mark = November 12-14; recapture = November 19-21), fall 2020 (mark = October 20-23; recapture = October 27-30); spring 2021 (mark = May 18-21). The recapture events were held just one week following the mark events to reduce opportunity for immigration/emigration of predators between events i.e., to avoid violation of a closed population for mark-recapture models.

We attempted a multiple of capture methods during the first year including raft electrofishing, fyke netting, snorkeling, and angling to reduce selectivity, both for species and size classes. Unfortunately, some methods proved ineffective during initial efforts. For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. And due to debris floating in the water column, there were concerns with fyke nets being pulled from their anchored locations and becoming entangled. Additionally, there were concerns with raft electrofishers and fyke nets incidentally capturing ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. Eventually, angling proved to be the most effective method for capturing piscine predators while minimizing potential impacts to steelhead but recognizing that angling can have both species and size selectivity (bias). From here on, we focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were allowed to fish anywhere within Deadwater Slough; periodically or upon filling a livewell anglers would boat or hike their catch to a processing station located at the boat ramp approximately 500 m downstream from the top of the slough. For each fish, processors would record the date of capture, species, total length (TL; mm), and whether the fish was previously marked. If not previously marked, each fish was provided a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique for each marking day. We also recorded whether marked individuals were released back to be captured again. For release, processors attempted to distribute released individuals throughout Deadwater Slough to facilitate mixing back into the population. We also recorded the crew (person or combination of persons) that captured each fish and the start and end times for each crew to estimate total effort.

@@

1. The population is closed (no immigration, emigration, births or deaths during the sampling period)
1. Marking a fish does not affect its chances of being caught
1. No loss of marks
1. No marks are missed or mistaken
1. All fish have the same chance of being caught in the second (and subsequent) sample(s)

Estimators fell into two broad categories: single census and multiple census. For the single census estimators, we treated the first week of sampling as the mark event, and the following (second) week as the recapture event, pooling data within each of those two weeks. Alternatively, the multiple census estimators treat each day as a survey, and use information about the total marked fish from all previous surveys to infer the total abundance.

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, due to logistical constraints, we were only able to sample during a single week in spring 2021 and released fish were left unmarked. Instead, to estimate the abundance of Northern Pikeminnow during spring 2021, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, and multiplied that ratio by the average abundance of Northern Pikeminnow from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our spring abundance estimate during the NRR Chinook Salmon emigration to be a conservative estimate of the number of Northern Pikeminnow present in the Deadwater Slough.

@@

Lastly, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during all field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon, other fishes (e.g., juvenile steelhead, juvenile Sockeye Salmon, Redside Shiner *Richardsonius balteatus*, etc.), and non-targets (e.g., macroinvertebrates, organic matter, etc.). Stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks immediately following lavage for later analysis in a controlled environment. For each sample, wet weight (grams) was recorded for the total combined stomach content as well as the contents consisting of fish parts/matter. Fish and fish remnants were identified to the lowest taxonomic unit, when possible, or were categorized as unknown. A subset of Northern Pikeminnow captures were euthanized for dissection after gastric lavage to validate the efficacy of the methodology.

@@

Gastric lavage was used in an attempt to document predation of juvenile Chinook Salmon by Northern Pikeminnow. However, we were ultimately interested in the potential total number or amount of juveniles consumed during the peaks of DSR (fall) and NRR (spring) emigrations. To estimate the consumption of juvenile Chinook Salmon in the Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight, converted from the average length of Northern Pikeminnow captured during our study and using the FSA package [@Ogle2021] in R. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

We ran three separate bioenergetic models for this study. For the first model, we used temperature data from a 78-day period of September 1 through November 17 where DSR emigrants are known to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin their downstream migration.<!-- Did we just use data from Lemhi here? All tributaries? Was this subjective? --> During this time, water temperatures typically exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon<!--citation?-->. This first model assumed no growth for Northern Pikeminnow. The second model was instead run for one full year, but also assumed no growth for Northern Pikeminnow. The assumption of zero growth was included to estimate consumption if either individual Northern Pikeminnow were not growing and/or the population is stable. Finally, the alternative third model shows how an increase of 10% body weight or population size changes the estimates of consumption over a full year period.

The model runs provided us with an estimate of the total biomass consumed by an individual Northern Pikeminnow for the given time period. We then multipled biomass consumed by an individual Northern Pikeminnow by the estimated population size for that time period to get an estimate of the total biomass consumed. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a eight of 12 g per individual. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.
